---
title: "Economic and Health Consequences of Severe Weather Events"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. The scope is to decide which events are most harmful. This report explores the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. 



## Data Processing

Libraries used
```{r results='hide', message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
```

```{r cache = TRUE}
if(!file.exists("projectdata.csv.bz2")){
  fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(fileUrl, destfile = "projectdata.csv.bz2", method = "curl")
}
projectdata <- read.csv("projectdata.csv.bz2")


```
### Health consequences
Summarize health problems(FATALITIES, INJURIES) by event type, removing events whith no consequences.
```{r}

sumFatalities <- projectdata %>%
            group_by(EVTYPE) %>%
            summarize(consequences = sum(FATALITIES)) %>%
            filter(consequences != 0) %>% 
            mutate(TYPE='FATALITY')


sumInjuries <- projectdata %>%
            group_by(EVTYPE) %>%
            summarize(consequences = sum(INJURIES)) %>%
            filter(consequences != 0) %>% 
            mutate(TYPE='INJURY')

healthProblems <- rbind(sumFatalities, sumInjuries)


```
How many events produce fatalities?

```{r}
nrow(sumFatalities)
```
How many events produce injuries?

```{r}
nrow(sumInjuries)
```
As there are too many events with consequences, only first 10 most devastating events are ploted.
```{r}
healthProblemsMost <- healthProblems %>%
            group_by(EVTYPE) %>%
            summarize(consequences = sum(consequences)) %>%
            arrange(-consequences) %>%
            top_n(10)
```
Select events by type, but only most imporant. Also, reorder events.
```{r}
healthProblemsMostByType <- healthProblems %>%
                            filter(EVTYPE %in% healthProblemsMost$EVTYPE)

healthProblemsMostByType$EVTYPE <- factor(healthProblemsMostByType$EVTYPE, levels = healthProblemsMost$EVTYPE[order(healthProblemsMost$consequences)])
```
### Economic consequences
Calculate real demage value
```{r}
projectdata <- projectdata %>%
  mutate(PropertyDamage = ifelse(toupper(PROPDMGEXP) =='K', yes = PROPDMG*1000, ifelse(toupper(PROPDMGEXP) =='M', PROPDMG*1000000, ifelse(toupper(PROPDMGEXP) == 'B', PROPDMG*1000000000, ifelse(toupper(PROPDMGEXP) == 'H', PROPDMG*100, PROPDMG))))) %>%
  mutate(CropDamage = ifelse(toupper(CROPDMGEXP) =='K', CROPDMG*1000, ifelse(toupper(CROPDMGEXP) =='M', CROPDMG*1000000, ifelse(toupper(CROPDMGEXP) == 'B', CROPDMG*1000000000, ifelse(toupper(CROPDMGEXP) == 'H', CROPDMG*100, CROPDMG)))))
```
Select events with most impact on economic situation of the communities. Reorder them so that most demaging events are shown.

```{r}

sumProp <- projectdata %>%
            group_by(EVTYPE) %>%
            summarize(consequences = sum(PropertyDamage)) %>%
            filter(consequences != 0) %>% 
            mutate(TYPE='PROP')


sumCrop <- projectdata %>%
            group_by(EVTYPE) %>%
            summarize(consequences = sum(CropDamage)) %>%
            filter(consequences != 0) %>% 
            mutate(TYPE='CROP')

economicProblems <- rbind(sumProp, sumCrop)



economicProblemsMost <- economicProblems %>%
            group_by(EVTYPE) %>%
            summarize(consequences = sum(consequences)) %>%
            arrange(-consequences) %>%
            top_n(10)

economicProblemsMost

economicProblemsMostByType <- economicProblems %>%
                            filter(EVTYPE %in% economicProblemsMost$EVTYPE)

economicProblemsMostByType

economicProblemsMostByType$EVTYPE <- factor(economicProblemsMostByType$EVTYPE, levels = economicProblemsMost$EVTYPE[order(economicProblemsMost$consequences)])


```
It can be seen that the weight of properties damage is bigger that the crop damage. Which events produce most damage for properties and which for crop?

```{r}
sumPropMax <- sumProp %>%
              arrange(-consequences) %>%
              top_n(10, consequences)

sumCropMax <- sumCrop %>%
              arrange(-consequences) %>%
              top_n(10, consequences)
sumCropMax

propCropMax <- rbind(sumPropMax, sumCropMax)

propCropMaxSum <- propCropMax %>%
            group_by(EVTYPE) %>%
            summarize(consequences = sum(consequences)) 

propCropMax$EVTYPE <- factor(propCropMax$EVTYPE, levels = propCropMaxSum$EVTYPE[order(propCropMaxSum$consequences)])
```

## Results

### Across the United States, which types of events are most harmful with respect to population health?

```{r}
 ggplot(data=healthProblemsMostByType, aes(x=EVTYPE, y=consequences, fill=TYPE)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(face="bold", color="#993333", size=12, angle=45)) + ylab("") + labs(title="Effect on population health")
```

### Across the United States, which types of events have the greatest economic consequences?



```{r}

 ggplot(data=economicProblemsMostByType, aes(y=consequences, x=EVTYPE, fill=TYPE)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(face="bold", color="#993333", size=12, angle=45)) + ylab("Value") + xlab("") + labs(title="Total economic consequences(properties and crop damages)")
```

```{r}

 ggplot(data=propCropMax, aes(y=consequences, x=EVTYPE, fill=TYPE)) + geom_bar(stat = "identity", position = "dodge") + theme(axis.text.x = element_text(face="bold", color="#993333", size=12, angle=45)) + ylab("Value") + xlab("") + labs(title="Most demaging events for properties and crops")
```

## Summary
**Tornadoes** are most demaging event in the US in terms of people health. In economic perspective, **floods** are most demaging events, followed by **hurricanes** and **tornadoes**. When looking deeply in the economic efects, **drought** can be highlited as most demaging event for crop.